The imprint of global magma oceans on exoplanet demographics

### 2023-03-13 src/tex/ms.tex
occurrence model caveats

% Define document class
%\documentclass[twocolumn]{aastex631}
\documentclass[modern]{aastex631}

@@

% Main body
\section{Introduction}
...just dropping material and references for now...
\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
\todo[inline]{``Points for the intro:
\bioverse\\
Transit survey to measure mass/radius of planets\\
Does the HZ have an inner edge?\\
Run-away GH\\
Magma oceans\\
Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman. Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
~\\
In contrast, PLATO and Nautilus are \\
Possibility of follow-up to measureme masses
''}

@@

\section{Baseline Model}\label{sec:met_baseline}
%The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
%Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
%To enable a quantitative assessment of this detectability, we injected the signal into a simulated planet sample before observing it with simulated surveys.
The goal of this study is to determine -- for different configurations of near-future exoplanet surveys -- the confidence level with which the runaway greenhouse threshold can be detected statistically. %in simulated exoplanet observations.
Our basic methodology was as follows:
We expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python. Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic samples of stars that host planets according to the observed exoplanet demographics.
We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).\todo[inline]{make sure rgh hypothesis is mentioned before, or rephrase}
\todo[inline]{\bioverse\ will need a short introduction before the modifications are mentioned. Probably should get a paragraph in the intro}
\begin{figure*}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/flowchart.pdf}
        \caption{Workflow of our hypothesis testing with \bioverse. First, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:met_rghmodel}.
        We then simulate an exoplanet survey, whereby selection effects and detection biases are introduced. Finally, we test the runaway greenhouse hypothesis based on data from the survey simulation.
        By iterating through these steps, we compute the statistical power of testing the hypothesis for different survey designs.}
        \label{fig:flowchart}
    \end{centering}
\end{figure*}
In this section, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, as well as the orbital parameters of the planets.

@@

\subsection{Synthetic planet sample}\label{sec:syn_planets}

Next, we assigned to the stellar sample planetary systems with frequencies, orbital parameters, and bulk properties derived from the \kepler\ mission.
We adopted the planetary occurrence rate model from \citet{Bergsten2022}, which defines the occurrence of small planets in radius and orbital period. Following \citet{Youdin2011a}, their inferred occurrence rate density can be expressed in the form:
\begin{equation}
    \frac{\mathrm{d}^2n}{\mathrm{d}R \, \partial P} = F_0 C_n g(R, P),
\end{equation}
where $F_0$ represents the average number of planets per star and $C_n$ is a normalization constant. The shape function $g(R,P)$ describing the occurrence distribution is the same as in \citet{Bergsten2022}. The functional form and stellar-mass dependent parameters (including a new fit to the sample of \kepler\ planets with hosts below $0.56\,M_{\odot}$) are included in \bioverse\ as the prominent planet-generation model; the original model based on SAG13 occurrence rates is preserved as a legacy feature.

With this updated distribution function, planets are generated as in the original \bioverse\ paper, but with an additional stellar mass dependence (for a star of a given mass, the relevant model parameters from the bin containing that mass are used). In summary, stars are assigned a number of planets in accordance with $F_0$, with planet parameters randomly drawn in proportion to the occurrence distribution function evaluated over a grid of radii and orbital periods.
% \todo[inline]{GALEN: Briefly describe underlying planet occurrence rates of synthetic population, its Mstar dependency, any caveats or assumptions.}
\todo[inline]{GALEN: The most relevant assumption of using our model is that, at the relevant fluxes (I think), we predict that super-Earths are pretty common. I think the "drop" in our model where super-Earths become less common happens at later fluxes than what is relevant in this paper... Let's double check this, and add a note in the first paragraph about it?.}

\subsection{Orbit parameters and planet masses}
Eccentric orbits alter the probability of a planet to transit \citep[e.g.,][]{Barnes2007a}.
The distribution of eccentricities $e$ of exoplanets has been found to resemble a Beta function \citep{Kipping2013b}, which we chose to draw synthetic eccentricities from.
Following~\citet{Kipping2013b}, we used a Beta distribution with parameters $a=0.867$ and $b=3.03$, and truncated the distribution at $e = 0.8$.
%Another parameter with an even higher impact on transit probabilities is the orbital inclination toward the observer $i$.
Assuming isotropic alignments of orbits, we assigned each planet an inclination drawn from a distribution uniform in $\cos(i)$.

@@

For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$, we assume an inflated radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013,Boukrouche2021}.
To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our non-runaway planets~\citep{Zeng2016}.
For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
This scenario does not take into account the expected additional radius change due to interiors that are molten due to greenhouse forcing~\citep{Bower2019}.
It further assumes a \textit{dry melt} without dissolved volatiles.

@@

\section{Exoplanet Survey Simulations and Hypothesis Testing}
The survey module of \bioverse\ converts the synthetic planet sample into a set of uncertainty-laden measurements on a subset of that sample.
This task includes selection of the targets, application of detection biases, and conducting simulated measurements, all of which are specific to the particular survey.
For each planet-level measurement such as radius or instellation, we draw the measured value from a normal distribution centered on the true value with a standard deviation set by the survey's precision.
We then follow a Bayesian hypothesis testing approach to assess various realizations of simulated surveys in terms of their ability to detect and characterize the runaway greenhouse transition.

@@

%    Current and near-future missions focusing on Earth-sized planets target several host star spectral types, and their exact sample selection is based on specific mission objectives.
%    We thus chose a broad selection function centered on general detectability of transits with state-of-the-art technology.
The runaway greenhouse effect becomes obsolete where no atmosphere can be maintained due to proximity to the host star and resulting atmospheric erosion.
Ensuring to stay well clear of such regions, we clear our sample from all planets with a net instellation $S > \SI{2000}{\watt\per\square\meter}$.
%\todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
We further consider only rocky planets with masses below \SI{2}{\Mearth}. % (Turbet+2020 covers only 0.1 to 2 Mearth)

@@

\subsection{Hypothesis tests}
We now turn to quantifying the ability of the simulated surveys to detect the runaway greenhouse transition and to constrain parameters of this trend.
To do this, we rely on a Bayesian hypothesis testing approach where we quantify the evidence of a hypothesis over another based on the (simulated) data.
For our specific problem, this implies comparing evidences for a demographic imprint of the runaway greenhouse effect to its absence.
\begin{note}
%    Bayesian model comparison requires that an alternative mode be specified against which the comparison is made.
\end{note}
    As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
    \begin{equation}
        H_0(\theta, S) = \theta,
    \end{equation}
    where $\theta$ is the set of parameters defining the radius distribution.
    We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
    As motivated above, this hypothesis takes the form of a step function in net instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
    Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
    The runaway greenhouse hypothesis is then defined as
\begin{equation}\label{eq:rgh_hypo}
    H_{\mathrm{rgh}}(\theta, S) =
        \begin{cases}
            H_0, &  S \leq S_\mathrm{thresh}\\
            \langle R_\mathrm{P}\rangle (f_\mathrm{rgh},\Delta R_\mathrm{stm}, \Delta R_\mathrm{wtr}), &  S > S_\mathrm{thresh}.
        \end{cases}
\end{equation}
    Here, $f_\mathrm{rgh}$ is the fraction of planets experiencing a runaway greenhouse effect.
    $\Delta R_\mathrm{stm}$ and $\Delta R_\mathrm{wtr}$ are predicted radius changes from the steam atmosphere and water incorporation models, respectively.
    They are assumed to act additively on the planet radii and thus on their average $\langle R_\mathrm{P}\rangle $.

@@

The measured radii $R_\mathrm{P, i}$ cannot be directly used for the hypothesis tests as they include intrinsic scatter that is not caused by measurement errors.
$H_{\mathrm{rgh}}$ and $H_0$ should thus be tested against a statistical estimator that represents the population mean.
To avoid binning and the artificial patterns it may introduce, we chose to test our hypotheses against a simple moving average $SMA$ along the instellation axis with a window of size \windowsize\ centered around each measurement.
We further computed the uncertainty of this moving average by propagating the individual measurement errors and applying a rolling standard error of the mean. \todo[inline]{specify error propagation}

@@

\section{Results}
\subsection{Statistical signature of the runaway greenhouse threshold}\label{sec:res_signature}
To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \var{wrr}$.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Synthetic planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable. Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, the radii of a fraction of planets are inflated by the amount indicated with gray lines. The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue). This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
Figure~\ref{fig:HnullHmo} shows the resulting planetary radii.
When ordered in orbital period space, the different planet types overlap, diluting the demographic imprint.
With net instellation as an independent variable, planets above and below the runaway greenhouse threshold separate: The runaway greenhouse-induced radius inflation introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
We also show the predictions of observable average planet radii from the statistical hypotheses defined above.
%The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
Within the runaway greenhouse regime, an average radius change of \avgRadiusChange\ occurs.

@@

\subsection{Detecting the runaway greenhouse transition with \plato}\label{sec:res_samplesize}
%\todo{Introduce survey simulation(s). PLATO, Cheops, (Ariel, LIFE/LIFE precursor, Nautilus)\\ What are threshold missions that are able to detect this signal? =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}

@@

Our finding that the expected trend is strong except for extremely low water mass fractions not only gives reason for optimism regarding its detection, it also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
%Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
%Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
Overall, searching for the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.

@@

\subsubsection{False positive scenarios}
\todo{what other processes could be confused with a magma ocean signal?\\
If there's no other subsubsection here, make this a (the final?) subsection in ``Discussion''.}
\begin{note}
    Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
    Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
    Both processes are being traded as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
%    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
    Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates.

    Another false positive contribution may come from potential unknown variations in the occurrence rate gradients in radius-instellation space.
    They can impact the statistical inference of the transitions, especially if these variations are similar to the expected pattern.
    Although an abrupt pattern at the expected runaway greenhouse transition seems unlikely, examples of steep occurrence rate density changes exist.
    An example is the "Neptune desert``, a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
    The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the instellation dependency of the runaway greenhouse transition.
    However, the Neptune desert occurs at smaller orbital periods and its location depends on a planet's size~\citep{Szabo2011}, which is not expected for the runaway greenhouse transition.
    \todo{is there overlap/confusion with the radius/density trend in \citet{Luque2022}?}
\end{note}

%\todo[inline]{Maps to Sect. \ref{sec:statpower_missions}.:}
%\begin{note}
%    \citet{Penny2019} mention that a significant increase in planet yield could be achieved if the telescope's slew speed was increased.
%\end{note}

\subsubsection{OPTIONAL...: Atmospheric signatures}
\todo[inline]{TIM: Ideas?\\
discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk)}
\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}

\subsection{Diagnostic power of near-future exoplanet missions}\label{sec:dis_samplesize}
\todo[inline]{discuss Nautilus}
The fidelity of a future detection or falsification of a runaway greenhouse signature will depend on the significance with which the null hypothesis can be excluded.
As a consequence, instrumentation and survey strategy aiming at probing the runaway greenhouse threshold should aim at maximizing the probability of a true positive detection.

@@

What other planned missions are suited to test the runaway greenhouse hypothesis?
The ongoing \cheops\ mission was designed as a follow-up mission to search for transits of planets discovered with other techniques, in particular with radial velocity measurements~\citep{Benz2021}.
As such, it will provide precise radius constraints on a sample of small planets; however, only a small number of planets with periods \SI{> 50}{\day} are being observed.
This largely limits \cheops' coverage to planets within the runaway greenhouse regime, preventing a detection of the transition.

As \cheops, \ariel~\citep{Puig2016} will be a follow-up mission that is not designed to provide a large number of new radius measurements.
\ariel's primary targets are larger planets in the range of sub-Neptune to Jupiter-like planets.
We thus do not expect a significant contribution to exploring the inner edge of the habitable zone for Earth-sized planets.

While not primarily designed to detect transiting planets, the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits and constrain their radii in the course of its mission~\citep{Montet2017}.
In a fraction of the cases, planets as small as \SI{2}{\Rearth} will be detected.
About 200 planets smaller than Neptune are expected to be found around M~dwarfs and thus may reach into the habitable zone.
The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
\todo[inline]{what about TESS?}

Taken together, \plato\ appears to be the most promising exoplanet mission for an empirical confirmation or falsification of the runaway greenhouse transition at this time.

>>>>

% Define document class
\documentclass[twocolumn]{aastex631}
%\documentclass[modern]{aastex631}

@@

% Main body
\section{Introduction}
...just dropping material and references for now...
\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}
\todo[inline]{``Points for the intro:
\bioverse\\
Transit survey to measure mass/radius of planets\\
Does the HZ have an inner edge?\\
Run-away GH\\
Magma oceans\\
Mention missions that are NOT statistical and therefore not considered here: CHEOPS, ARIEL, Roman. Other missions/concepts do not directly measure radii and not considered here: Habitable Worlds Observatory, LIFE.\\
~\\
In contrast, PLATO and \nautilus\ are \\
Possibility of follow-up to measureme masses
''}

@@

\section{Baseline Model}\label{sec:met_baseline}
%The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
%Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
%To enable a quantitative assessment of this detectability, we injected the signal into a simulated planet sample before observing it with simulated surveys.
The goal of this study is to determine -- for different configurations of near-future exoplanet surveys -- the confidence level with which the runaway greenhouse threshold can be detected statistically. %in simulated exoplanet observations.
Our basic methodology was as follows:
We expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python. Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic samples of stars that host planets according to the observed exoplanet demographics.
We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).\todo[inline]{make sure rgh hypothesis is mentioned before, or rephrase}
\todo[inline]{\bioverse\ will need a short introduction before the modifications are mentioned. Probably should get a paragraph in the intro}
\begin{figure*}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/flowchart.pdf}
        \caption{Workflow of our hypothesis testing with \bioverse. First, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:met_rghmodel}.
        We then simulate an exoplanet survey, whereby selection effects and detection biases are introduced. Finally, we test the runaway greenhouse hypothesis based on data from the survey simulation.
        By iterating through these steps, we compute the statistical power of testing the hypothesis for different survey designs.}
        \label{fig:flowchart}
    \end{centering}
\end{figure*}
In this section, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, as well as the orbital parameters of the planets.
An overview of our key assumptions and model parameters can be found in Table~\ref{tab:params_table}.

@@

\subsection{Synthetic planet sample}\label{sec:syn_planets}
Next, we assigned to the stellar sample planetary systems with frequencies, orbital parameters, and bulk properties derived from the \kepler\ mission.
We adopted the planetary occurrence rate model from~\citet{Bergsten2022}, which defines the occurrence of small planets in radius and orbital period. Following \citet{Youdin2011a}, their inferred occurrence rate density can be expressed in the form:
\begin{equation}
    \frac{\mathrm{d}^2n}{\mathrm{d}R \, \partial P} = F_0 C_n g(R, P),
\end{equation}
where $F_0$ represents the average number of planets per star and $C_n$ is a normalization constant. The shape function $g(R,P)$ describing the occurrence distribution is the same as in \citet{Bergsten2022}. The functional form and stellar-mass dependent parameters (including a new fit to the sample of \kepler\ planets with hosts below $0.56\,M_{\odot}$) are included in \bioverse\ as the prominent planet-generation model; the original model based on SAG13 occurrence rates is preserved as a legacy feature.

With this updated distribution function, planets are generated as in the original \bioverse\ paper, but with an additional stellar mass dependence (for a star of a given mass, the relevant model parameters from the bin containing that mass are used). In summary, stars are assigned a number of planets in accordance with $F_0$, with planet parameters randomly drawn in proportion to the occurrence distribution function evaluated over a grid of radii and orbital periods.
\todo[inline]{GALEN: The most relevant assumption of using our model is that, at the relevant fluxes (I think), we predict that super-Earths are pretty common. I think the "drop" in our model where super-Earths become less common happens at later fluxes than what is relevant in this paper... Let's double check this, and add a note in the first paragraph about it?.}

\subsection{Orbit parameters and planet masses}\label{sec:met-orbits_masses}
Eccentric orbits alter the probability of a planet to transit \citep[e.g.,][]{Barnes2007a}.
The distribution of eccentricities $e$ of exoplanets has been found to resemble a Beta function \citep{Kipping2013b}, which we chose to draw synthetic eccentricities from.
Following~\citet{Kipping2013b}, we used a Beta distribution with parameters $a=0.867$ and $b=3.03$, and truncated the distribution at $e = 0.8$.
%Another parameter with an even higher impact on transit probabilities is the orbital inclination toward the observer $i$.
Assuming isotropic alignments of orbits, we assigned each planet an inclination drawn from a distribution uniform in $\cos(i)$.

@@

For all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Abe1988,Kasting1988,Nakajima1992,Goldblatt2013,Leconte2013,Boukrouche2021}.
To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
Their calculations rely on the same mass-radius relations for rocky interiors that we apply for our non-runaway planets~\citep{Zeng2016}.
For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
This scenario does not take into account the expected additional radius change due to interiors that are molten due to greenhouse forcing~\citep{Bower2019}.
It further assumes a \textit{dry melt} without dissolved volatiles.

@@

\section{Exoplanet Survey Simulations and Hypothesis Testing}\label{sec:met_surveys-hypotests}
The survey module of \bioverse\ converts the synthetic planet sample into a set of uncertainty-laden measurements on a subset of that sample.
This task includes selection of the targets, application of detection biases, and conducting simulated measurements, all of which are specific to the particular survey.
For each planet-level measurement such as radius or instellation, we draw the measured value from a normal distribution centered on the true value with a standard deviation set by the survey's precision.
We then follow a Bayesian hypothesis testing approach to assess various realizations of simulated surveys in terms of their ability to detect and characterize the runaway greenhouse transition.

@@

%    Current and near-future missions focusing on Earth-sized planets target several host star spectral types, and their exact sample selection is based on specific mission objectives.
%    We thus chose a broad selection function centered on general detectability of transits with state-of-the-art technology.
The runaway greenhouse effect becomes obsolete both for very small instellations and where no atmosphere can be maintained due to proximity to the host star and resulting atmospheric erosion.
Ensuring to stay well clear of such regions, we clear our sample from all planets with a net instellation $S < \SI{10}{\watt\per\square\meter}$ or $S > \SI{2000}{\watt\per\square\meter}$.
%\todo[inline]{use criterion of Zahnle \& Caitling 2017 instead? $0.8 S^{0.25} < R$}
We further consider only rocky planets with masses below \SI{2}{\Mearth}. % (Turbet+2020 covers only 0.1 to 2 Mearth)

@@

\subsection{Hypothesis tests}
We now turn to quantifying the ability of the simulated surveys to detect the runaway greenhouse transition and to constrain parameters of this trend.
To do this, we rely on a Bayesian hypothesis testing approach where we quantify the evidence of a hypothesis over another based on the (simulated) data.
For our specific problem, this implies comparing evidences for a demographic imprint of the runaway greenhouse effect to its absence.
%\begin{note}
%    Bayesian model comparison requires that an alternative mode be specified against which the comparison is made.
%\end{note}
    As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
    \begin{equation}
        H_0(\theta, S) = \theta,
    \end{equation}
    where $\theta$ is the set of parameters defining the radius distribution.
    We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
    As motivated above, this hypothesis takes the form of a step function in net instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
    Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
    The runaway greenhouse hypothesis is then defined as
\begin{equation}\label{eq:rgh_hypo}
    H_{\mathrm{rgh}}(\theta, S) =
        \begin{cases}
            H_0, &  S \leq S_\mathrm{thresh}\\
            \langle R_\mathrm{P}\rangle (f_\mathrm{rgh},\Delta R_\mathrm{stm}, \Delta R_\mathrm{wtr}), &  S > S_\mathrm{thresh}.
        \end{cases}
\end{equation}
    Here, $f_\mathrm{rgh}$ is the fraction of planets experiencing a runaway greenhouse effect.
    $\Delta R_\mathrm{stm}$ and $\Delta R_\mathrm{wtr}$ are predicted radius changes from the steam atmosphere and water incorporation models, respectively.
    They are assumed to act additively on the planet radii and thus on their average $\langle R_\mathrm{P}\rangle $.

@@

The measured radii $R_\mathrm{P, i}$ cannot be directly used for the hypothesis tests as they include intrinsic scatter that is not caused by measurement errors.
$H_{\mathrm{rgh}}$ and $H_0$ should thus be tested against a statistical estimator that represents the population mean.
To avoid binning and the artificial patterns it may introduce, we chose to test our hypotheses against a simple moving average $SMA$ along the instellation axis with a window of size \windowsize\ centered around each measurement.\footnote{See Appendix~\ref{app:binnedstats} for a robustness test using a different estimator.}
We further computed the uncertainty of this moving average by propagating the individual measurement errors and applying a rolling standard error of the mean. \todo[inline]{specify error propagation}

@@

% include table with key parameters
\input{table_parameters}

\section{Results}
\subsection{Statistical signature of the runaway greenhouse threshold}\label{sec:res_signature}
To characterize the population-level imprint of individual radius changes, we generated a generic planet population with an injected runaway greenhouse effect assuming a water fraction $x_\mathrm{H_2O} = \var{wrr}$.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Synthetic planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of net instellation. Only the planets marked as transiting are observable.
            Above the runaway greenhouse threshold $S_\mathrm{thresh} = \var{injectedSthresh}$, some planets maintain their original radii (green crosses) whereas some have their radii inflated (blue crosses) by the amount indicated with gray lines.
        The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue line).
            This runaway greenhouse hypothesis can be tested against the null hypothesis $H_0$ (green line), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
Figure~\ref{fig:HnullHmo} shows the resulting planetary radii.
When ordered in orbital period space, the different planet types overlap, diluting the demographic imprint.
With net instellation as an independent variable, planets above and below the runaway greenhouse threshold separate: The runaway greenhouse-induced radius inflation introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
We also show the predictions of observable average planet radii from the statistical hypotheses defined above.
%The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
Within the runaway greenhouse regime, an average radius change of \avgRadiusChange\ occurs.

@@

\subsection{Detecting the runaway greenhouse transition with \plato}\label{sec:res_samplesize}
%\todo{Introduce survey simulation(s). PLATO, Cheops, (Ariel, LIFE/LIFE precursor, \nautilus\)\\ What are threshold missions that are able to detect this signal? =>X meter working for Y years, or X' meter working for Y' years => conclusion could be that there is interesting science to do with intermediate size telescope.}

@@

Our finding that the expected trend is strong except for extremely low water mass fractions gives reason for optimism regarding its detection.
It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
Such constraints will shed light on whether the initial water content is variable~\citep{Lichtenberg2019} or there is a predominant quantity.
Ultimately, such measurements will shed light on the question of if planets like Earth received their water from primordial sources during their formation or if it was supplied later~\citep[e.g.,][]{Raymond2004}.
%Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
%Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
Overall, searching for the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.

@@

\subsubsection{False positive scenarios}\label{sec:dis_falsepositive}
Runaway greenhouse climates are not the only physical mechanism that may cause a change in transit radius for a subset of planets.
Some of the most prominent alternative scenarios involve atmospheric loss due to either photoevaporation through high-energy radiation by the host star~\citep[e.g.,][]{Owen2013,Jin2014,Mordasini2020a} or due to residual heat from the planet's interior shortly after formation~\citep{Ginzburg2016b,Ginzburg2018,Gupta2019}.
Both processes are being traded as potentially sculpting the observed bimodality in the radius distribution of small exoplanets~\citep{Fulton2017,VanEylen2018}.
%    Like the magma ocean effect discussed here, these processes reduce the radii of some planets, leading to a decrease in average measured planet radii in a specific region of the planetary parameter space.
Either of them leads to a decrease of planet radius for planets close to their host star, which is distinct from the radius inflation introduced by runaway greenhouse climates.

Another false positive contribution may come from potential unknown variations in the occurrence rate gradients in radius-instellation space.
They can impact the statistical inference of the transitions, especially if these variations are similar to the expected pattern.
Although an abrupt pattern at the expected runaway greenhouse transition seems unlikely, examples of steep occurrence rate density changes exist.
An example is the ``Neptune desert'', a triangular region of low planet occurrence density of close-in planets in period-radius space~\citep{Szabo2011,Mazeh2016,Dreizler2020b}.
The shape of this region is such that smaller planets become less frequent the closer to the star they are, which to some degree resembles the pattern introduced by the instellation dependency of the runaway greenhouse transition.
However, the Neptune desert occurs at smaller orbital periods and its location depends on a planet's size~\citep{Szabo2011}, which is not expected for the runaway greenhouse transition.
\todo{is there overlap/confusion with the radius/density trend in \citet{Luque2022}?}

\subsubsection{OPTIONAL...: Atmospheric signatures}\label{sec:dis_atmospheres}
\todo[inline]{TIM: Ideas?\\
discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk)}
\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}

\subsection{Diagnostic power of near-future exoplanet missions}\label{sec:dis_samplesize}
The fidelity of a future detection or falsification of a runaway greenhouse signature will depend on the significance with which the null hypothesis can be excluded.
As a consequence, instrumentation and survey strategy aiming at probing the runaway greenhouse threshold should aim at maximizing the probability of a true positive detection.

@@

What other planned missions are suited to test the runaway greenhouse hypothesis?
\kepler\ and \ktwo\ have contributed a large number of discovered terrestrial-sized planets, but few of them are in the habitable zone and their host stars are typically too faint for RV follow-up~\citep{Dressing2015}.

Similarly, the Transiting Exoplanet Survey Satellite~\citep[\tess,][]{Ricker2014a} planet sample lacks temperate, small planets around bright host stars~\citep{Ment2023}, as was expected from planet yield calculations~\citep{Barclay2018}.

The ongoing \cheops\ mission was designed as a follow-up mission to search for transits of planets discovered with other techniques, in particular with radial velocity measurements~\citep{Benz2021}.
As such, it will provide precise radius constraints on a sample of small planets; however, only a small number of planets with periods \SI{> 50}{\day} are being observed.
This largely limits \cheops' coverage to planets within the runaway greenhouse regime, preventing a detection of the transition.

As \cheops, the Atmospheric Remote sensing Infrared Exoplanet Large survey~\citep[\ariel,][]{Puig2016} will be a follow-up mission that is not designed to provide a large number of new radius measurements.
\ariel's primary targets are larger planets in the range of sub-Neptune to Jupiter-like planets.
We thus do not expect a significant contribution to statistically exploring the inner edge of the habitable zone for Earth-sized planets.

While not primarily designed to detect transiting planets, the \rst~\citep{Spergel2015} is expected to yield $\sim 10^5$ transiting planets on short orbits and constrain their radii in the course of its mission~\citep{Montet2017}.
In a fraction of the cases, planets as small as \SI{2}{\Rearth} will be detected.
About 200 planets smaller than Neptune are expected to be found around M~dwarfs and thus may reach into the habitable zone.
The \rst\ could thus provide a useful sample to explore the runaway greenhouse transition, albeit with a predominant focus on water-rich (sub-)Neptunes~\citep[e.g.,][]{Pierrehumbert2022}.
